---
title: "Law School Ideological Bias"
author: "Kirthi Chigurupati"
description: "Gov 50 Final Project"
output:
  distill::distill_article:
    self_contained: false
---

https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/8WYLCG

## Introduction

With organizations like the Federalist Society increasing in political power, many Americans are left feeling that the branch of our government meant to balance the power of the others is becoming more biased. Recent supreme court rulings, for instance, have been split among the political party of the president that appointed them. There is argument that through the further ideological bias present in prestigous law school curriculums, a bias is being taught to students that affect them further into their careers as judges. My question is whether prestigous law school experiences are causing further ideological bias in their students. 


## Data and research design

In my project, I plan to explore whether prestigious law school attendance can create ideological bias in judges. My explanatory variable, therefore, will be the whether the nominee attended a private institution for college. 

My data was collected by Charles S. Bullock III, Richard L. Vining Jr., Ethan D. Boldt and tracks judicial appointments from  Georgia between 1991 to 2014 on various political factors, qualification, and ideological bias to find the relationship between political donations and likely hood of appointment, coming to a claim of causality between those variables. The research design is observational and correlation analysis. The analysis covers the appointments by 3 different Governors from both political parties and 4,835 unique nominees. For the purposes of my research, I will be analyzing the data on law school details. 


```{r include=FALSE}
#Options
options(
  scipen = 9,
  width = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center")
library(tidyverse)
library(ggplot2)
library(broom)

appointments <- read.csv("Politics-Interim-Judicial-Appts.csv")

```


```{r}
appointments |>
  group_by(year) |>
  summarize (polarity = mean(imputed_gov_nom_ideo_distance)) |>
  ggplot (aes(x=year, y=polarity)) +
  geom_smooth (color = "#900000") +
  labs (
    x = "Year", 
    y = "Ideological polarity of the Appointee (in percentage)", 
    title = "Trend of polarity in judicial appointments"
  )
```


As seen in the above trend, there is a significant increase in the polarity of the candidates overtime. Between 1990 and 1995 there is an apparent downward trend, but around 1994, the trend suddenly and sharply turns. After noting this pattern, I decided to hold year as a fixed constant to ensure it was not affecting the results.


```{r}
#data

appointments |>
  ggplot(mapping = aes(x=imputed_gov_nom_ideo_distance)) +
  geom_histogram() +
  labs (x = "Nominal Ideological Distance",
        y = "Count", 
        title = "Distribution of Outcome Variable")

```

The above graph shows the distribution of the outcome variable — **Nominal Ideological Distance**. The variable is tracked in the absolute value, showing the distance from ideologically pure of each of the appointees. It is on a scale to 3, as is consistent with the tracking method: **Bonica’s CFScore**.


```{r}
appointments |>
  ggplot(mapping = aes(x=median_lsat)) +
  geom_histogram() +
  labs (x = "Median LSAT Score",
        y = "Count", 
        title = "Distribution of Explanatory Variable")
```


## Results 

The bivariate relationship between the explanatory variable and the outcome variable is visualized as a scatter plot below. 

```{r}
appointments |>
  ggplot(mapping = aes(x=median_lsat, y=gov_nom_ideo_distance)) +
  geom_point(shape = 19, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1)

```
The bivariate plot suggests that there is no relationship between a higher median lsat score for their law school towards ideological impurity. The regression model below will offer more insight about this relationship. 

## Regression Models and Results

```{r}
mod_1 <- lm(
    data=appointments,
  imputed_gov_nom_ideo_distance ~ median_lsat + as.factor(year))


summary(
  mod_1
  )$coefficients[1:2,] |>
  knitr::kable(digits = 2)

```

The bivariate regression model has a coefficent of 0 and a p-value of 0.01. If we set alpha = 0.05, that means our finding is statistically significant. The bivariate regression is therefore consistent with the scatter plot findings.

```{r}

mod_2 <- lm(
    data=appointments,
  imputed_gov_nom_ideo_distance ~ 
    instatel + pubpvt + median_lsat + as.factor(year))


summary(
  mod_2
  )$coefficients[1:4,] |>
  knitr::kable(digits = 2)



```
The table above shows the results from a multivariate model that includes the following cofounders related to law school: 

 - whether various judicial appointees attended an in-state school (instatel)
 - whether they attended a private institution (pubpvt)
 - year as a fixed effect as explained above
 
The regression results show that there is 0 correlation between the variables, and that this result is statistically significant and we can reject the null (p-value = 0). 

In both bivariate and multivariate models, the association between median LSAT score of a judge's law school and their ideological bias is statistically significant at the test level α=0.05. The p-value on the coefficient in both models is 0. The coefficient's p-value of 0 suggests an almost 0% probability that the estimated association observed is as meaningless and extreme as the estimated association from the null distribution. There is a near 100% probability that the association is not from random chance. 


Assuming the sampling distribution is approximately normal, a confidence interval of 95% can be constructed with the alpha of 0.05. 

# Causality and cofounders. 

Since the explanatory variable of ideological bias was not randomly assigned, it is not a "treatment" variable. This means a lack of casual relationship cannot be assumed, rather, there is no correlation between the variables. 

The multivariate model attempts to control for whether the law school attended is private or public. This could be a possible cofounder because median LSAT does not necessarily reflect the prestige of an institution, rather the lack of oversight and reputation of being a private institution could cause it. Likewise, having an out of state college, which was also accounted for in the multivariate model, could create some correlation between geographic traits in creating bias. 

Other cofounders could exits, but are not fully available in the data. For instance, undergrad degree location. If there is bias in the law school teaching process, undergraduate curriculums too should be analyzed. Exposure to varying socioeconomic status can also be a cofounder, as it can impact a judge's ability to make ideologically pure decisions. 

## Conclusion

The regression demonstrates that there is no relationship between the prestige of a law school and how ideologically biased the judges come out of them may be. There is a statistical significance. THe multivariate model that controlled for other cofounders about the law school, including whether it is instate or if it was private, suggested the same and was also statistically significant. 

The analysis is limited in that it only draws from the appointment of one state. Having other regional trials would have increased the accuracy. Likewise, Not all forms of education are accounted for (undergrad and secondary level). This is likely because the data was not collected to draw the same conclusions about schooling, it was collected to analyze political contributions. Possible improvements are expanding the scope of the study and understanding more about the various elements of prestigious education that may compel ideoglogical bias like legacy status, income level, and regional bias. 

